---
title: "EV-biomarker study"
author: Alvin Meltsov
date: "Compiled on `r format(Sys.time(), '%d/%m/%y')`"
output:
  rmdformats::robobook:
    number_sections: true
    code_folding: hide
    fig_width: 8
    fig_height: 6
---

# Prelude: Setup environment

## Load data and environment

Setup the environment and load dependencies, defined in external `back.R` script:

1. Load the dependencies, set up multithreading.
2. Set color groups and values.
3. Define convenience functions for preprocessing and visualisation.
4. Load and transform phenotype data and the data matrix.
5. Query BioMart for gene annotation and alternative names.
6. Load also important gene lists.

```{r warning = F, message = F}
## !RUNTHIS

## load the project dependencies
source("./scripts/load_deps.R")

## load the model metrics functions
source("./scripts/metrics_functions.R")

## check if theres a R data object already, othewise run the data muncher
# and save the preprocessed data into the RData
data_subfolder <- "./data/orig/"
dataobj <- paste0(data_subfolder, "ev_data.RData")
if (file.exists(dataobj)) {
  load(dataobj)
} else {
  ## load in the backend data muncher
  commandArgs <- function(...)
    list(data_subfolder = data_subfolder)
  dir.create(data_subfolder, showWarnings = F)
  invisible(source("./scripts/de_runner.R"))
  save.image(file = dataobj)
}

```

In this study we are working with a set of 26 samples over proliferative and 3 different secretory phases:

```{r warning = F, message = F}

barplotter <- function(phenotype_obj, title_text = "Training samples")
  phenotype_obj %>%
    ggplot(aes(cyclephase, fill = group)) +
    geom_bar() +
    theme_thesis() +
    xlab("Phases") +
    ylab("Number of samples") +
    ggtitle(title_text) +
    theme(plot.title = element_text()) +
    scale_fill_manual(values = status_grouping)

pheno %>% barplotter | pheno_test %>% barplotter(title = "Testing samples")

```

## Look at general clustering of samples {.tabset}

This step is performed to just look at any possible deviant samples that are systematically different from rest of the samples.

### PCA

First 3 components over all the genes.

```{r warning = F, message = F}

plot_ly() %>%
  add_trace(
    data = global_expr_pca,
    x = ~ PC1,
    y = ~ PC2,
    z = ~ PC3,
    type = "scatter3d",
    text = ~ paste0(samplename),
    color = ~ cyclephase,
    colors = color_grouping_alt,
    symbol = ~ group,
    symbols = c("circle", "circle-open"),
    marker = list(size = 6),
    mode = "markers"
  )

```

### UMAP

But for large and poorly normalised data we are better off using the UMAP algrithm.

```{r warning = F, message = F}

plot_ly() %>%
  add_trace(
    data = global_expr_umap,
    x = ~ UMAP1,
    y = ~ UMAP2,
    type = "scatter",
    text = ~ paste0(samplename),
    color = ~ cyclephase,
    colors = color_grouping_alt,
    symbol = ~ group,
    symbols = c("circle", "circle-open"),
    marker = list(size = 6),
    mode = "markers"
  )

```

### Correlation heatmap

```{r warning = F, message = F}

cor_ht <-
  expr_tpm %>%
  select(-c("cyclephase", "group")) %>%
  column_to_rownames("samplename") %>%
  as.matrix %>%
  t %>%
  cor(method = "spearman") %>%
  heatmaply_cor(
    limits = c(0, 1),
    col_side_colors = expr_mat %>% select(where(is.character)) %>% column_to_rownames("samplename"),
    row_side_colors = expr_mat %>% select(where(is.character)) %>% column_to_rownames("samplename")
  )

cor_ht

```

From these preliminary clusterings we can see that the HUT23 paired sample seems to be switched and some samples seem to be outliers based on global expression profile alone. Lets switch or remove them and rerun the analysis.

```{r warning = F, message = F}
## !RUNTHIS

## check if theres a R data object already, othewise run the data muncher 
# and save the preprocessed data into the RData
rm(list = ls())

## reload the project dependencies
source("./scripts/load_deps.R")

## reload the model metrics functions
source("./scripts/metrics_functions.R")

data_subfolder <- "./data/filtered/"
dataobj <- paste0(data_subfolder, "ev_data_filt.RData")
if (file.exists(dataobj)) {
  load(dataobj)
} else {
  ## load in the backend data muncher with arguments
  commandArgs <- function(...)
    list(
      switch_sample = "HUT23",
      remove_sample = c("HUT71_UF", "HUT10_UF", "HUT17_UF", "HUT1_UF"),
      data_subfolder = data_subfolder
      ) #  c("HUT71_UF", "HUT1_UF", "HUT01_UF", "HUT17_UF", "HUT10_UF"))
  dir.create(data_subfolder, showWarnings = F)
  invisible(source("./scripts/de_runner.R"))
  save.image(file = dataobj)
}

```

### Barplot - adjusted

```{r warning = F, message = F}

barplotter <- function(phenotype_obj, title = "Training samples")
  phenotype_obj %>%
    ggplot(aes(cyclephase, fill = group)) +
    geom_bar() +
    theme_thesis() +
    xlab("Phases") +
    ylab("Number of samples") +
    ggtitle("Training samples") +
    theme(plot.title = element_text()) +
    scale_fill_manual(values = status_grouping)

pheno %>% barplotter | pheno_test %>% barplotter(title = "Testing samples")

```

### PCA - adjusted

First 3 components over all the genes.

```{r warning = F, message = F}

plot_ly() %>%
  add_trace(
    data = global_expr_pca,
    x = ~ PC1,
    y = ~ PC2,
    z = ~ PC3,
    type = "scatter3d",
    text = ~ paste0(samplename),
    color = ~ cyclephase,
    colors = color_grouping_alt,
    symbol = ~ group,
    symbols = c("circle", "circle-open"),
    marker = list(size = 6),
    mode = "markers"
  )

```

### UMAP - adjusted

But for large and poorly normalised data we are better off using the UMAP algrithm.

```{r warning = F, message = F}

plot_ly() %>%
  add_trace(
    data = global_expr_umap,
    x = ~ UMAP1,
    y = ~ UMAP2,
    type = "scatter",
    text = ~ paste0(samplename),
    color = ~ cyclephase,
    colors = color_grouping_alt,
    symbol = ~ group,
    symbols = c("circle", "circle-open"),
    marker = list(size = 6),
    mode = "markers"
  )

```

### Correlation heatmap - adjusted

```{r warning = F, message = F}

cor_ht <-
  expr_tpm %>%
  select(-c("cyclephase", "group")) %>%
  column_to_rownames("samplename") %>%
  as.matrix %>%
  t %>%
  cor(method = "spearman") %>%
  heatmaply_cor(
    limits = c(0, 1),
    col_side_colors = expr_mat %>% select(where(is.character)) %>% column_to_rownames("samplename"),
    row_side_colors = expr_mat %>% select(where(is.character)) %>% column_to_rownames("samplename")
  )

cor_ht

```


## Run DESeq2, edgeR and limma-voom {.tabset}

Here we run standard differentation analysis toolsets DESeq2 and limma-voom for all the samples and all the comparisons. Therefore, we look at differences between successive mens cycle phase:

* UF:
  * Proliferative vs LH2/3
  * LH2/3 vs LH7/8
  * LH7/8 vs LH11/13
* Biopsy:
  * Proliferative vs LH2/3
  * LH2/3 vs LH7/8
  * LH7/8 vs LH11/13

... and every comparison between the UF and Biopsy within the same cycle group (ie UF-LH2/3 vs Biopsy-LH2/3).

### UF {.tabset}

Venn diagram of DE genes and differential analysis results tables of DE genes for the **UF** samples:

```{r warning = F, message = F}
## !RUNTHIS

uf_out_join <-
  unique(c(
    deseq_res_phase$UF$pro_vs_pre$locus,
    deseq_res_phase$UF$pre_vs_rec$locus,
    deseq_res_phase$UF$rec_vs_post$locus,
    voom_res_phase$UF$pro_vs_pre$locus,
    voom_res_phase$UF$pre_vs_rec$locus,
    voom_res_phase$UF$rec_vs_post$locus
  )) %>% {
    tibble(
      locus = .,
      deseq_pro_vs_pre = . %in% deseq_res_phase$UF$pro_vs_pre$locus,
      deseq_pre_vs_rec = . %in% deseq_res_phase$UF$pre_vs_rec$locus,
      deseq_rec_vs_post = . %in% deseq_res_phase$UF$rec_vs_post$locus,
      voom_pro_vs_pre = . %in% voom_res_phase$UF$pro_vs_pre$locus,
      voom_pre_vs_rec = . %in% voom_res_phase$UF$pre_vs_rec$locus,
      voom_rec_vs_post = . %in% voom_res_phase$UF$rec_vs_post$locus
      )
    }

p1 <-
  uf_out_join %>%
  filter(if_any(contains("deseq_"))) %>%
  ggplot(aes(A = deseq_pro_vs_pre, B = deseq_pre_vs_rec, C = deseq_rec_vs_post)) +
  geom_venn(
    show_percentage = F, fill_alpha = 0.3,
    set_names = c("Pro vs. LH2", "LH2 vs LH7", "LH7 vs LH12")
  ) + theme_thesis_venn() + ggtitle("DESeq2 significant DE genes")

p2 <-
  uf_out_join %>%
  filter(if_any(contains("voom_"))) %>%
  ggplot(aes(A = voom_pro_vs_pre, B = voom_pre_vs_rec, C = voom_rec_vs_post)) +
  geom_venn(
    show_percentage = F, fill_alpha = 0.3,
    set_names = c("Pro vs. LH2", "LH2 vs LH7", "LH7 vs LH12")
  ) + theme_thesis_venn() + ggtitle("voom significant DE genes")

p1 | p2

uf_out_join %>% downtable(filterable = T, searchable = T, defaultPageSize = 5)

```

Collect the DE different genes from **DESeq2** and **limma-voom**:

#### DESeq2: {.tabset}

##### Proliferative vs Pre-receptive
```{r warning = F, message = F}
## sadly reactable does not work in loops...
deseq_res_phase$UF[[1]] %>%
  rename("{names(deseq_res_phase$UF[1])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Pre-receptive vs Receptive
```{r warning = F, message = F}
deseq_res_phase$UF[[2]] %>%
  rename("{names(deseq_res_phase$UF[2])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Receptive vs Post-receptive
```{r warning = F, message = F}
deseq_res_phase$UF[[3]] %>%
  rename("{names(deseq_res_phase$UF[3])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```

#### limma-voom: {.tabset}

##### Proliferative vs Pre-receptive
```{r warning = F, message = F}
## sadly reactable does not work in loops...
voom_res_phase$UF[[1]] %>%
  rename("{names(voom_res_phase$UF[1])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Pre-receptive vs Receptive
```{r warning = F, message = F}
voom_res_phase$UF[[2]] %>%
  rename("{names(voom_res_phase$UF[2])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Receptive vs Post-receptive
```{r warning = F, message = F}
voom_res_phase$UF[[3]] %>%
  rename("{names(voom_res_phase$UF[3])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```


### Biopsy {.tabset}

Venn diagram of DE genes and differential analysis results tables of DE genes for the **biopsy** samples:

```{r warning = F, message = F}
## !RUNTHIS

bio_out_join <-
  unique(c(
    deseq_res_phase$biopsy$pro_vs_pre$locus,
    deseq_res_phase$biopsy$pre_vs_rec$locus,
    deseq_res_phase$biopsy$rec_vs_post$locus,
    voom_res_phase$biopsy$pro_vs_pre$locus,
    voom_res_phase$biopsy$pre_vs_rec$locus,
    voom_res_phase$biopsy$rec_vs_post$locus
  )) %>% {
    tibble(
      locus = .,
      deseq_pro_vs_pre = . %in% deseq_res_phase$biopsy$pro_vs_pre$locus,
      deseq_pre_vs_rec = . %in% deseq_res_phase$biopsy$pre_vs_rec$locus,
      deseq_rec_vs_post = . %in% deseq_res_phase$biopsy$rec_vs_post$locus,
      voom_pro_vs_pre = . %in% voom_res_phase$biopsy$pro_vs_pre$locus,
      voom_pre_vs_rec = . %in% voom_res_phase$biopsy$pre_vs_rec$locus,
      voom_rec_vs_post = . %in% voom_res_phase$biopsy$rec_vs_post$locus
      )
    }

p1 <-
  bio_out_join %>%
  filter(if_any(contains("deseq_"))) %>%
  ggplot(aes(A = deseq_pro_vs_pre, B = deseq_pre_vs_rec, C = deseq_rec_vs_post)) +
  geom_venn(
    show_percentage = F, fill_alpha = 0.3,
    set_names = c("Pro vs. LH2", "LH2 vs LH7", "LH7 vs LH12")
  ) + theme_thesis_venn() + ggtitle("DESeq2 significant DE genes")

p2 <-
  bio_out_join %>%
  filter(if_any(contains("voom_"))) %>%
  ggplot(aes(A = voom_pro_vs_pre, B = voom_pre_vs_rec, C = voom_rec_vs_post)) +
  geom_venn(
    show_percentage = F, fill_alpha = 0.3,
    set_names = c("Pro vs. LH2", "LH2 vs LH7", "LH7 vs LH12")
  ) + theme_thesis_venn() + ggtitle("voom significant DE genes")

p1 | p2

bio_out_join %>% downtable(filterable = T, searchable = T, defaultPageSize = 5)

```

Collect the DE different genes from **DESeq2** and **limma-voom**:

#### DESeq2: {.tabset}

##### Proliferative vs Pre-receptive
```{r warning = F, message = F}
## sadly reactable does not work in loops...
deseq_res_phase$biopsy[[1]] %>%
  rename("{names(deseq_res_phase$biopsy[1])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Pre-receptive vs Receptive
```{r warning = F, message = F}
deseq_res_phase$biopsy[[2]] %>%
  rename("{names(deseq_res_phase$biopsy[2])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Receptive vs Post-receptive
```{r warning = F, message = F}
deseq_res_phase$biopsy[[3]] %>%
  rename("{names(deseq_res_phase$biopsy[3])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```

#### limma-voom: {.tabset}

##### Proliferative vs Pre-receptive
```{r warning = F, message = F}
## sadly reactable does not work in loops...
voom_res_phase$biopsy[[1]] %>%
  rename("{names(voom_res_phase$biopsy[1])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Pre-receptive vs Receptive
```{r warning = F, message = F}
voom_res_phase$biopsy[[2]] %>%
  rename("{names(voom_res_phase$biopsy[2])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Receptive vs Post-receptive
```{r warning = F, message = F}
voom_res_phase$biopsy[[3]] %>%
  rename("{names(voom_res_phase$biopsy[3])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```

### UF vs biopsy {.tabset}

Venn diagram of DE genes and differential analysis results tables of DE genes for the **biopsy** samples in one receptivity group when compared to matching receptivity group in **UF** set:

```{r warning = F, message = F}
## !RUNTHIS

out_join <-
  unique(c(
    deseq_res_group$pro$UF_vs_biopsy$locus,
    deseq_res_group$pre$UF_vs_biopsy$locus,
    deseq_res_group$rec$UF_vs_biopsy$locus,
    deseq_res_group$post$UF_vs_biopsy$locus,
    voom_res_group$pro$UF_vs_biopsy$locus,
    voom_res_group$pre$UF_vs_biopsy$locus,
    voom_res_group$rec$UF_vs_biopsy$locus,
    voom_res_group$post$UF_vs_biopsy$locus
  )) %>% {
    tibble(
      locus = .,
      deseq_pro = . %in% deseq_res_group$pro$UF_vs_biopsy$locus,
      deseq_pre = . %in% deseq_res_group$pre$UF_vs_biopsy$locus,
      deseq_rec = . %in% deseq_res_group$rec$UF_vs_biopsy$locus,
      deseq_post = . %in% deseq_res_group$post$UF_vs_biopsy$locus,
      voom_pro = . %in% voom_res_group$pre$UF_vs_biopsy$locus,
      voom_pre = . %in% voom_res_group$pre$UF_vs_biopsy$locus,
      voom_rec = . %in% voom_res_group$rec$UF_vs_biopsy$locus,
      voom_post = . %in% voom_res_group$post$UF_vs_biopsy$locus
      )
    }

p1 <-
  out_join %>%
  filter(if_any(contains("deseq_"))) %>%
  ggplot(aes(A = deseq_pro, B = deseq_pre, C = deseq_rec, D = deseq_post)) +
  geom_venn(
    show_percentage = F, fill_alpha = 0.3,
    set_names = c("Prolif", "LH2", "LH7", "LH12")
  ) + theme_thesis_venn() + ggtitle("DESeq2 significant DE genes")

p2 <-
  out_join %>%
  filter(if_any(contains("voom_"))) %>%
  ggplot(aes(A = voom_pro, B = voom_pre, C = voom_rec, D = voom_post)) +
  geom_venn(
    show_percentage = F, fill_alpha = 0.3,
    set_names = c("Prolif", "LH2", "LH7", "LH12")
  ) + theme_thesis_venn() + ggtitle("voom significant DE genes")

p1 | p2

out_join %>% downtable(filterable = T, searchable = T, defaultPageSize = 5)

```

Collect the DE different genes from **DESeq2** and **limma-voom**:

#### DESeq2: {.tabset}

##### Proliferative - UF vs Bio
```{r warning = F, message = F}
## sadly reactable does not work in loops...
deseq_res_group[[1]]$UF_vs_biopsy %>%
  rename("{names(deseq_res_group[1])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Pre-receptive - UF vs Bio
```{r warning = F, message = F}
deseq_res_group[[2]]$UF_vs_biopsy %>%
  rename("{names(deseq_res_group[2])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Receptive - UF vs Bio
```{r warning = F, message = F}
deseq_res_group[[3]]$UF_vs_biopsy %>%
  rename("{names(deseq_res_group[3])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Post-receptive - UF vs Bio
```{r warning = F, message = F}
deseq_res_group[[4]]$UF_vs_biopsy %>%
  rename("{names(deseq_res_group[4])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```

#### limma-voom: {.tabset}

##### Proliferative - UF vs Bio
```{r warning = F, message = F}
## sadly reactable does not work in loops...
voom_res_group[[1]]$UF_vs_biopsy %>%
  rename("{names(voom_res_group[1])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Pre-receptive - UF vs Bio
```{r warning = F, message = F}
voom_res_group[[2]]$UF_vs_biopsy %>%
  rename("{names(voom_res_group[2])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Receptive - UF vs Bio
```{r warning = F, message = F}
voom_res_group[[3]]$UF_vs_biopsy %>%
  rename("{names(voom_res_group[3])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```
##### Post-receptive - UF vs Bio
```{r warning = F, message = F}
voom_res_group[[4]]$UF_vs_biopsy %>%
  rename("{names(voom_res_group[4])}" := "locus") %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)
```

## TODO: Check the reference samples

The data was sequenced in two batches, with the exception of sample HUT71. This same biopsy sample was sequenced in both batches, allowing us to have a control sample to assess the similiarity of batches, and in the case of bias, use it to correct for batch effects.

# TODO: Are UF-biopsy paired samples similar?

# Other genesets {.tabset}

Here we apply the beREADY model for the counts. Due to large number of samples having bad coverage of the housekeeper genes, we apply here a slight modification to the geometric mean normalisation step. Namely, we use shifted logarithm instead of the natural logarithm found in the geometric mean calculation. Therefore these scores may not be completely comparable with the reference samples.

For other genesets, we use TPM normalised counts.

## beREADY {.tabset}

```{r warning=F, message=F}

# use this function to apply the geometric mean on the counts
geom_norm_col <- function(
    dat,
    vals_from = "molecule_count",
    finders = c("CYC1", "HMBS", "TBP", "SDHA"),
    .rename = F
    ) {
  dat %>%
    select(-where(is.numeric) | one_of(vals_from)) %>%
    pivot_wider(
      names_from = c("sample", "batch"),
      names_sep = ">",
      values_from = all_of(vals_from)
    ) %>%
    mutate_if(is.numeric, function(x) {
      x / exp(mean(log(x[.$locus %in% finders])))
    }) %>%
    pivot_longer(where(is.numeric),
      names_to = c("sample", "batch"),
      names_sep = ">",
      values_to = ifelse(.rename, vals_from, "norm_count")
    )
}

# replace ENSID's with the common names
beready_ensids <-
  annot %>%
  filter(external_gene_name %in% c(beready_genes, finders)) %>%
  select(-description)

expr_dat <-
  expr_mat %>%
    pivot_longer(
      where(is_bare_numeric),
      names_to = "ensembl_gene_id",
      values_to = "molecule_count"
    ) %>%
    right_join(
      beready_ensids,
      by = "ensembl_gene_id"
    ) %>%
    select(-ensembl_gene_id) %>%
    # eliminate any duplicates from the name harmonisatoin
    distinct(samplename, external_gene_name, .keep_all = T)

norm_test_dat <-
  expr_dat %>%
    rename(
      locus = external_gene_name,
      sample = samplename,
      batch = group
    ) %>%
    select(-cyclephase) %>%
    # apply the shifted logarithm geom. normalisation
    geom_norm_col %>%
    pivot_wider(names_from = "locus", values_from = "norm_count")

## TMP
model_obj <-
  silent_beready_model(
    tacseqApp::controls_data$`READY NC controls`,
    tacseqApp::targets_data$`Dynamic filtering geneset`,
    norm_test_dat
  )

```


### Shifted lognorm counts

Here we subset the genes in the beREADY assay, do shifted log-normalisation and check how the expression rates differ between samplescheck how the expression rates differ between samples.

```{r warning = F, message = F, fig.height = 12}

expr_dat %>%
  filter(external_gene_name %in% beready_genes) %>%
  distinct(.keep_all = T) %>%
  ggplot(aes(log2(molecule_count + 1), external_gene_name, fill = group)) +
  geom_boxplot() +
  theme_thesis() +
  xlab("") +
  ylab("Molecule counts log2") +
  scale_fill_manual(values = status_grouping) +
  scale_y_discrete(limits = rev)

```

As one can see, sample have a pretty similar expressional profile, therefore they shouldnt really differ in receptivity testing too.

### PCA with reference samples

First two principal components of the classes with the beREADY reference sampleset. The shapes here represent the original sample class and the colors the beREADY class predictions.

```{r warning = F, message = F}

shape_grouping <- c("LH2"  = 6, "LH7" = 5, "LH12" = 2, "Prolif" = 4)
add_coloring <- c(
  "LH2" = color_grouping_alt[["pre"]],
  "LH7" = color_grouping_alt[["rec"]],
  "LH12" = color_grouping_alt[["post"]],
  "Prolif" = color_grouping_alt[["pro"]]
  )

## make the PCA object for later on
expr_pca <-
  model_obj %$%
  { join_model_pca(pca_all, score_table) } %>%
  left_join(
    pheno %>% select(-group), by = c("sample" = "samplename")
  ) %>%
  mutate(group = case_when(
    cyclephase == "pro" ~ "Prolif",
    cyclephase == "pre" ~ "LH2",
    cyclephase == "rec" ~ "LH7",
    cyclephase == "post" ~ "LH12",
    T ~ group
  )) %>%
  ## exclude no-result smales
  filter(label != "No result")

pca_w_fields <- function(pca_obj, shapes, coloring)
  pca_obj %>%
  filter(group != "Reference") %>%
  ggplot(aes(PC1, PC2, label = sample, color = label, shape = group)) +
  stat_density2d(
    data = pca_obj %>% filter(group == "Reference"),
    mapping = aes(x = PC1, y = PC2, fill = label, alpha = after_stat(level)),
    geom = "polygon",
    bins = 5,
    contour_var = "ndensity",
    inherit.aes = F,
    show.legend = F
  ) +
  geom_point(size = 3, stroke = 1) +
  labs(color = "") +
  scale_color_manual(values = coloring) +
  scale_alpha_continuous(limits = c(0, 20), breaks = seq(0, 10, by = 2)) +
  scale_shape_manual(values = shapes) +
  # coord_cartesian(xlim = c(NA, 6)) +
  theme_thesis() +
  theme(
    legend.position = c(.85, .15),
    # panel.grid.minor = element_line(),
    legend.direction = "vertical",
    legend.text.align = 0,
    # legend.box.just = 0,
    legend.box = "vertical"
 )

```

Plot it for both UF and for biopsy separate:

```{r warning = F, message = F}

uf_pca <- pca_w_fields(
  expr_pca %>% filter(str_detect(sample, "UF") | group == "Reference"),
  shape_grouping,
  c(color_grouping, add_coloring)
)

bio_pca <- pca_w_fields(
  expr_pca %>% filter(str_detect(sample, "biopsy") | group == "Reference"),
  shape_grouping,
  c(color_grouping, add_coloring)
)

(uf_pca | bio_pca) +
  plot_annotation(
    title = "PCA with beREADY reference samples",
    subtitle = "
      PCA of UF (A) and biopsy (B) samples with beREADY reference samples.\n
      Colors represent predicted classes, shapes the original cyclephase of the sample.",
    tag_levels = "A",
    theme = theme_thesis_venn()
    )

```

And add the results table and an interactive PCA table. #TODO add accuracy metrics and the reference receptivity groups.

```{r warning = F, message = F}

model_obj$score_table %>%
  downtable(sortable = T, filterable = T)

inner_join(model_obj$score_table, pheno %>% filter(group == "UF"),  by = c("Sample Name"="samplename")) %>%
  mutate(matched = case_when(
    `Predicted Group` == "Pre-receptive" & cyclephase == "pre" ~ T,
    `Predicted Group` == "Late-Receptive" & cyclephase == "rec" ~ T,
    `Predicted Group` == "Receptive" & cyclephase == "rec" ~ T,
    `Predicted Group` == "Early-Receptive" & cyclephase == "rec" ~ T,
    `Predicted Group` == "Post-receptive" & cyclephase == "post" ~ T,
    `Predicted Group` == "Pre-receptive" & cyclephase == "pro" ~ T,
    T ~ F
  )) %$% 
  matched %>%
  mean

```

```{r warning = F, message = F}

plot_ly() %>%
  add_trace(
    data = expr_pca %>%
      mutate(
        type = ifelse(is.na(dataset), "reference", dataset),
        cyclephase = ifelse(is.na(cyclephase), label, group)
        ),
    x = ~ PC1,
    y = ~ PC2,
    type = "scatter",
    text = ~ paste0(sample,
      "<br>Receptivity group: ", cyclephase, 
      "<br>Predicted group: ", label,
      "<br>Dataset: ", dataset),
    color = ~ group, 
    symbol = ~ type,
    symbols = c("circle", "circle-open"),
    marker = list(size = 6),
    mode = "markers"
  )

```

## TED genes {.tabset}

TED set consist of 73 genes that are found retrospectively via tuning ML models on previously analysed biomarker study datasets.

```{r warning=F, message=F}

ted_ensids <-
  annot %>%
  filter(external_gene_name %in% ted_genes) %>%
  select(-description)

filt_counts <-
  expr_tpm %>%
    pivot_longer(
      where(is_bare_numeric),
      names_to = "ensembl_gene_id",
      values_to = "molecule_count"
    ) %>%
    right_join(
      ted_ensids,
      by = "ensembl_gene_id"
    ) %>%
    select(-ensembl_gene_id) %>%
    # eliminate any duplicates from the name harmonisatoin
    distinct(samplename, external_gene_name, .keep_all = T)

```

### Boxplot

```{r warning = F, message = F, fig.height=16}

filt_counts %>%
  distinct(.keep_all = T) %>%
  ggplot(aes(log2(molecule_count), external_gene_name, fill = group)) +
  geom_boxplot() +
  theme_thesis() +
  xlab("") +
  ylab("Molecule counts log2") +
  scale_fill_manual(values = status_grouping) +
  scale_y_discrete(limits = rev)

# ggsave("./figures/box_ted.svg", width = 6, height = 9)

```

### PCA

```{r warning = F, message = F}

filt_counts_wide <-
  filt_counts %>%
  pivot_wider(
    id_cols = "samplename",
    names_from = "external_gene_name",
    values_from = "molecule_count"
    ) %>%
  mutate(across(where(is.numeric), ~replace(., is.na(.), 0))) %>%
  filter(if_any(where(is.numeric), ~ sum(.) > 10)) %>%
  rowwise %>%
  filter(if_any(where(is.numeric), ~ sum(.) > 10)) %>%
  ungroup

expr_pca <-
  filt_counts_wide %>%
  recipe %>%
  update_role("samplename", new_role = "id") %>%
  step_zv(all_numeric()) %>%
  step_normalize(all_numeric()) %>%
  step_pca(all_numeric(), num_comp = 3) %>%
  prep(strings_as_factors = F) %>%
  bake(new_data = filt_counts_wide) %>%
  left_join(pheno, by = "samplename") %>%
  filter(!is.na(samplename))

plot_ly() %>%
  add_trace(
    data = expr_pca,
    x = ~ PC1,
    y = ~ PC2,
    z = ~ PC3,
    type = "scatter3d",
    text = ~ paste0(samplename, "<br>Receptivity group:", cyclephase),
    color = ~ cyclephase,
    colors = color_grouping_alt,
    symbol = ~ group,
    symbols = c("circle", "circle-open"),
    marker = list(size = 6),
    mode = "markers"
  )

```

### UMAP

```{r warning = F, message = F}

expr_umap <-
  filt_counts_wide %>%
  recipe %>%
  update_role("samplename", new_role = "id") %>%
  step_zv(all_numeric()) %>%
  step_normalize(all_numeric()) %>%
  step_umap(all_numeric(), num_comp = 3) %>%
  prep(strings_as_factors = F) %>%
  bake(new_data = filt_counts_wide) %>%
  left_join(pheno, by = "samplename") %>%
  filter(!is.na(samplename))

plot_ly() %>%
  add_trace(
    data = expr_umap,
    x = ~ UMAP1,
    y = ~ UMAP2,
    z = ~ UMAP3,
    type = "scatter3d",
    text = ~ paste0(samplename, "<br>Receptivity group:", cyclephase),
    color = ~ cyclephase,
    colors = color_grouping_alt,
    symbol = ~ group,
    symbols = c("circle", "circle-open"),
    marker = list(size = 6),
    mode = "markers"
  )


```


## ERA genes {.tabset}

```{r warning=F, message=F}

era_ensids <-
  annot %>%
  filter(external_gene_name %in% era_genes) %>%
  select(-description)

filt_counts <-
  expr_tpm %>%
    pivot_longer(
      where(is_bare_numeric),
      names_to = "ensembl_gene_id",
      values_to = "molecule_count"
    ) %>%
    right_join(
      era_ensids,
      by = "ensembl_gene_id"
    ) %>%
    select(-ensembl_gene_id) %>%
    # eliminate any duplicates from the name harmonisatoin
    distinct(samplename, external_gene_name, .keep_all = T)

```

### Boxplot

There are >200 genes, so this is not gonna make sense...

```{r warning = F, message = F, fig.height=24}

filt_counts %>%
  distinct(.keep_all = T) %>%
  ggplot(aes(log2(molecule_count + 1), external_gene_name, fill = group)) +
  geom_boxplot() +
  theme_thesis() +
  xlab("") +
  ylab("Molecule counts log2") +
  scale_fill_manual(values = status_grouping) +
  scale_y_discrete(limits = rev)

# ggsave("./figures/box_ted.svg", width = 6, height = 9)

```

### PCA

```{r warning = F, message = F}

filt_counts_wide <-
  filt_counts %>%
  pivot_wider(
    id_cols = "samplename",
    names_from = "external_gene_name",
    values_from = "molecule_count"
    ) %>%
  mutate(across(where(is.numeric), ~replace(., is.na(.), 0))) %>%
  filter(if_any(where(is.numeric), ~ sum(.) > 10)) %>%
  rowwise %>%
  filter(if_any(where(is.numeric), ~ sum(.) > 10)) %>%
  ungroup

expr_pca <-
  filt_counts_wide %>%
  recipe %>%
  update_role("samplename", new_role = "id") %>%
  step_zv(all_numeric()) %>%
  step_normalize(all_numeric()) %>%
  step_pca(all_numeric(), num_comp = 3) %>%
  prep(strings_as_factors = F) %>%
  bake(new_data = filt_counts_wide) %>%
  left_join(pheno, by = "samplename") %>%
  filter(!is.na(samplename))

plot_ly() %>%
  add_trace(
    data = expr_pca,
    x = ~ PC1,
    y = ~ PC2,
    z = ~ PC3,
    type = "scatter3d",
    text = ~ paste0(samplename, "<br>Receptivity group:", cyclephase),
    color = ~ cyclephase,
    colors = color_grouping_alt,
    symbol = ~ group,
    symbols = c("circle", "circle-open"),
    marker = list(size = 6),
    mode = "markers"
  )

```


### UMAP

```{r warning = F, message = F}

expr_umap <-
  filt_counts_wide %>%
  recipe %>%
  update_role("samplename", new_role = "id") %>%
  step_zv(all_numeric()) %>%
  step_normalize(all_numeric()) %>%
  step_umap(all_numeric(), num_comp = 3) %>%
  prep(strings_as_factors = F) %>%
  bake(new_data = filt_counts_wide) %>%
  left_join(pheno, by = "samplename") %>%
  filter(!is.na(samplename))

plot_ly() %>%
  add_trace(
    data = expr_umap,
    x = ~ UMAP1,
    y = ~ UMAP2,
    z = ~ UMAP3,
    type = "scatter3d",
    text = ~ paste0(samplename, "<br>Receptivity group:", cyclephase),
    color = ~ cyclephase,
    colors = color_grouping_alt,
    symbol = ~ group,
    symbols = c("circle", "circle-open"),
    marker = list(size = 6),
    mode = "markers"
  )

```

# Biomarkers for UF samples

## Clustering with the geneset from DE results

Here we create a possible set of biomarkers from the DESeq2 DE comparisons between cycle types and/or sample types. We define two sets here:

1. UF only biomarkers, with the following set equation:

$$ F_{UF} = (UF_{pro-vs-pre} \cup UF_{pre-vs-rec}) \cap (UF_{pre-vs-rec} \cup UF_{rec-vs-post}) \\ \cap (UF_{rec_vs_post} \cup UF_{pro-vs-pre}) $$

2. Biopsy + UF biomarkers, with the following set equation, given that we have found the set $F_{biopsy}$ in the same way as :

$$ F_{UF} \cup F_{biopsy} $$

Therefore the biopsy + UF biomarker set is a subset of the UF_biomarker set. This set is found in a efficient manner of converting boolean values to numerical and filtering out genes with more than 1 truth values.

The list of genes with annotations are here for the UF only biomarkers:

```{r warning = F, message = F}
## !RUNTHIS

uf_genes <-
  uf_out_join %>%
  select(c("locus", contains("deseq_"))) %>%
  rowwise %>%
  mutate(sumlogic = sum(c_across(where(is.logical)), na.rm = T)) %>%
  filter(sumlogic >= 2) %$%
  locus

annot %>%
  filter(ensembl_gene_id %in% uf_genes) %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)


bio_genes <-
  bio_out_join %>%
  select(c("locus", contains("deseq_"))) %>%
  rowwise %>%
  mutate(sumlogic = sum(c_across(where(is.logical)), na.rm = T)) %>%
  filter((sumlogic >= 2) & (locus %in% uf_genes)) %$%
  locus

annot %>%
  filter(ensembl_gene_id %in% bio_genes) %>%
  downtable(filterable = T, searchable = T, defaultPageSize = 5)

```

Following clustering algorithms are first "trained" on just UF samples, and then all samples are projected onto it.

### UF set {.tabset}

```{r warning=F, message=F}

filt_counts <-
  expr_tpm %>%
    pivot_longer(
      where(is_bare_numeric),
      names_to = "ensembl_gene_id",
      values_to = "molecule_count"
    ) %>%
    filter(ensembl_gene_id %in% uf_genes & group == "UF")

```


#### PCA

```{r warning = F, message = F}

filt_counts_wide <-
  filt_counts %>%
  pivot_wider(
    id_cols = "samplename",
    names_from = "ensembl_gene_id",
    values_from = "molecule_count"
    ) %>%
  mutate(across(where(is.numeric), ~replace(., is.na(.), 0))) %>%
  filter(if_any(where(is.numeric), ~ sum(.) > 10)) %>%
  rowwise %>%
  filter(if_any(where(is.numeric), ~ sum(.) > 10)) %>%
  ungroup

expr_pca <-
  filt_counts_wide %>%
  recipe %>%
  update_role("samplename", new_role = "id") %>%
  step_zv(all_numeric()) %>%
  step_normalize(all_numeric()) %>%
  step_pca(all_numeric(), num_comp = 3) %>%
  prep(strings_as_factors = F) %>%
  bake(new_data = filt_counts_wide) %>%
  left_join(pheno, by = "samplename") %>%
  filter(!is.na(samplename))

plot_ly() %>%
  add_trace(
    data = expr_pca,
    x = ~ PC1,
    y = ~ PC2,
    z = ~ PC3,
    type = "scatter3d",
    text = ~ paste0(samplename, "<br>Receptivity group:", cyclephase),
    color = ~ cyclephase,
    colors = color_grouping_alt,
    symbol = ~ group,
    symbols = c("circle", "circle-open"),
    marker = list(size = 6),
    mode = "markers"
  )

```

#### UMAP

```{r warning = F, message = F}

expr_umap <-
  filt_counts_wide %>%
  recipe %>%
  update_role("samplename", new_role = "id") %>%
  step_zv(all_numeric()) %>%
  step_normalize(all_numeric()) %>%
  step_umap(all_numeric(), num_comp = 3) %>%
  prep(strings_as_factors = F) %>%
  bake(new_data = filt_counts_wide) %>%
  left_join(pheno, by = "samplename") %>%
  filter(!is.na(samplename))

plot_ly() %>%
  add_trace(
    data = expr_umap,
    x = ~ UMAP1,
    y = ~ UMAP2,
    z = ~ UMAP3,
    type = "scatter3d",
    text = ~ paste0(samplename, "<br>Receptivity group:", cyclephase),
    color = ~ cyclephase,
    colors = color_grouping_alt,
    symbol = ~ group,
    symbols = c("circle", "circle-open"),
    marker = list(size = 6),
    mode = "markers"
  )

```

### UF + biopsy set {.tabset}

```{r warning=F, message=F}

filt_counts <-
  expr_tpm %>%
    pivot_longer(
      where(is_bare_numeric),
      names_to = "ensembl_gene_id",
      values_to = "molecule_count"
    ) %>%
    filter(ensembl_gene_id %in% bio_genes)

```

#### PCA

```{r warning = F, message = F}

filt_counts_wide <-
  filt_counts %>%
  pivot_wider(
    id_cols = "samplename",
    names_from = "ensembl_gene_id",
    values_from = "molecule_count"
    ) %>%
  mutate(across(where(is.numeric), ~replace(., is.na(.), 0))) %>%
  filter(if_any(where(is.numeric), ~ sum(.) > 10)) %>%
  rowwise %>%
  filter(if_any(where(is.numeric), ~ sum(.) > 10)) %>%
  ungroup

expr_pca <-
  filt_counts_wide %>%
  filter(samplename %in% (pheno %>% filter(group == "UF") %$% samplename)) %>%
  recipe %>%
  update_role("samplename", new_role = "id") %>%
  step_zv(all_numeric()) %>%
  step_normalize(all_numeric()) %>%
  step_pca(all_numeric(), num_comp = 3) %>%
  prep(strings_as_factors = F) %>%
  bake(new_data = filt_counts_wide) %>%
  left_join(pheno, by = "samplename") %>%
  filter(!is.na(samplename))

plot_ly() %>%
  add_trace(
    data = expr_pca,
    x = ~ PC1,
    y = ~ PC2,
    z = ~ PC3,
    type = "scatter3d",
    text = ~ paste0(samplename, "<br>Receptivity group:", cyclephase),
    color = ~ cyclephase,
    colors = color_grouping_alt,
    symbol = ~ group,
    symbols = c("circle", "circle-open"),
    marker = list(size = 6),
    mode = "markers"
  )

```

#### UMAP

```{r warning = F, message = F}

expr_umap <-
  filt_counts_wide %>%
  filter(samplename %in% (pheno %>% filter(group == "UF") %$% samplename)) %>%
  recipe %>%
  update_role("samplename", new_role = "id") %>%
  step_zv(all_numeric()) %>%
  step_normalize(all_numeric()) %>%
  step_umap(all_numeric(), num_comp = 3) %>%
  prep(strings_as_factors = F) %>%
  bake(new_data = filt_counts_wide) %>%
  left_join(pheno, by = "samplename") %>%
  filter(!is.na(samplename))

plot_ly() %>%
  add_trace(
    data = expr_umap,
    x = ~ UMAP1,
    y = ~ UMAP2,
    z = ~ UMAP3,
    type = "scatter3d",
    text = ~ paste0(samplename, "<br>Receptivity group:", cyclephase),
    color = ~ cyclephase,
    colors = color_grouping_alt,
    symbol = ~ group,
    symbols = c("circle", "circle-open"),
    marker = list(size = 6),
    mode = "markers"
  )

```

## New housekeepers

Here we read in the housekeepers from the [HRT database](https://housekeeping.unicamp.br/?homePageHuman), also described in this [article](https://academic.oup.com/nar/article/49/D1/D947/5871367). Then, we check if those genes are non-differential in any comparison between the receptivity groups and if they are highly expressed in UF samples. After all that we select **four** least differential, most expressed genes based on our dataset.

The parameters used on the website were:

* The **Human** atlas
* The **Uterus** dataset
* At least log RPKM expression rate of **69** (so that there are roughly 100 genes included with high expression rates)

```{r warning = F, message = F}
## !RUNTHIS

# load in
hs_set <- read_csv("./data_raw/hrt_uterus_rpkm69.csv")

# show the table of housekeepers
hs_set %>% downtable(filterable = T)

# transform it to match local data
hs_set %<>%
  ## select only rank column as others are not informative
  select(c("Rank", "Gene Symbol", "Ensembl ID", "Normalized Expression")) %>%
  rename(
    ensembl_transcript_id = `Ensembl ID`,
    external_gene_name = `Gene Symbol`,
    hrt_rank = Rank,
    norm_mean_expr = `Normalized Expression`
  )

```

We would like to get housekeepers that are well expressed, but at the same time not too highly expressed so that it wouldnt' most of the reaction mixture.

Let's see how those genes express in both our data and in the reference as mean. Here we calculate TMM RPKM values for the genes with the `edgeR` package as the reference values were also normalised in that way.

Then we'll exclude those genes that were in any of the DE genes group.

```{r warning = F, message = F}
## !RUNTHIS

gene_mat <-
  raw_mat %>%
  select(c("gene_id", contains("UF"))) %>%
  rename(ensembl_gene_id = gene_id) %>%
  geneid_converter(annot) %>%
  drop_na(external_gene_name) %>%
  mutate(across(where(is.numeric), ~replace(., is.na(.), 0)))

edger_gene_len <-
  gene_len %>%
  filter(gene_id %in% gene_mat$external_gene_name) %>%
  # match the order of the gene matrix
  arrange(
    gene_id,
    gene_mat %>%
      filter(external_gene_name %in% gene_len$gene_id) %$%
      external_gene_name
    ) %$%
  len

norm_rpkm_mat <-
  gene_mat %>%
  column_to_rownames("external_gene_name") %>%
  edgeR::DGEList() %>%
  edgeR::calcNormFactors() %>%
  edgeR::rpkm(gene.length = edger_gene_len) %>%
  as_tibble(rownames = "external_gene_name") %>%
  # and finally, filter out genes of interest for less comp overhead
  filter(external_gene_name %in% hs_set$external_gene_name)

norm_rpkm_mat %<>%
  ## now calculate the mean per gene, the sds and the f-scores
  mutate(
    external_gene_name = external_gene_name,
    mean_rpkm =
      rowMeans(across(where(is_bare_numeric)), na.rm = T),
    sds_rpkm =
      matrixStats::rowSds(as.matrix(across(where(is_bare_numeric)), na.rm = T)),
    # do the ANOVA for every gene and return the F-stat
    fstat_rpkm =
      across(everything()) %>%
      pivot_longer(where(is_bare_numeric), names_to = "samplename", values_to = "rpkms") %>%
      inner_join(pheno %>% select(-c("group", "dataset")), by = "samplename") %>%
      group_by(external_gene_name) %>%
      # ... and run anova with the cyclephase variables in mind
      group_map(~ tidy(aov(rpkms ~ cyclephase, data = .x))$statistic[1]) %>%
      unlist,
    .keep = "unused"
  )

norm_rpkm_mat %<>%
  ## and add the housekeeper atlas table, the intersect just to be sure
  inner_join(
    hs_set %>% select(c("norm_mean_expr", "external_gene_name")),
    by = "external_gene_name"
    ) %$%
  bind_rows(
    "HRT Atlas" = tibble(external_gene_name, norm_mean_expr),
    "HUT set" = tibble(external_gene_name, mean_rpkm, sds_rpkm, fstat_rpkm),
    .id = "dataset"
  ) %>%
  # harmonise the norm mean columns of atlas and our set
  mutate(norm_mean_expr = ifelse(is.na(mean_rpkm), norm_mean_expr, mean_rpkm)) %>%
  select(-mean_rpkm)

```

Potential housekeepers **before** filtering out with DE genes: `r dim(norm_rpkm_mat)[1]`

```{r warning = F, message = F}
## !RUNTHIS

# now exclude genes in any of the DE group
de_genes <-
  annot %>%
    filter(
      ensembl_gene_id %in%
        c(uf_out_join$locus, bio_out_join$locus) #, out_join$locus)
    ) %$%
    external_gene_name

norm_rpkm_mat %<>% filter(!(external_gene_name %in% de_genes))

```

Potential housekeepers **after** filtering out with DE genes: `r dim(norm_rpkm_mat)[1]`

```{r warning = F, message = F}

# do a side-by-side violin plot of the expression rate of housekeepers in reference and in our data
breaks <- 10^(-10:10)
minor_breaks <- rep(1:9, 21)*(10^rep(-10:10, each=9))
norm_rpkm_mat %>%
  ggplot(aes(dataset, norm_mean_expr)) +
    geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
    theme_thesis() +
    xlab("Datasets") +
    ylab("TMM RPKM values") +
    scale_y_log10(breaks = breaks, minor_breaks = minor_breaks) + 
    theme(
      panel.grid.minor = element_line(),
      panel.grid.major = element_line()
    )

```

Although comparable, and luckily not heavily skewed towards zero, the mean expressions of those genes do vary between whats expected of the HRT dataset and what we observe

Lets see how many genes we get in the lower quartile region of the HRT geneset, and the second quartile of our dataset. Additionally, lets order the genes also by the F-statistic values to see the most stably expressed genes of the bunch.

Genes listed in **bold** are the top 4 selected for housekeepers.

```{r warning = F, message = F}
## !RUNTHIS

# get the .25-.5 quantile range of genes in both datasets
uf_hs <-
  norm_rpkm_mat %>%
  group_by(dataset) %>%
  mutate(
    lower_quart = quantile(norm_mean_expr, .25),
    upper_quart = quantile(norm_mean_expr, .75)
  ) %>%
  ungroup %>%
  filter(
    # dont include the Atlas filtering, as we only care about the expression of HUT set
    # ( dataset == "HRT Atlas" & norm_mean_expr < lower_quart ) |
      ( dataset == "HUT set" & norm_mean_expr < upper_quart & norm_mean_expr > lower_quart )
  ) %>%
  select(-contains("quart"))

# add ranking from the HRT dataset
uf_hs %<>%
  left_join(
    hs_set %>% select(c("hrt_rank", "external_gene_name")) %>% distinct(external_gene_name, .keep_all = T),
    by = "external_gene_name"
  ) %>%
  # remove the atlas dataset now to remove duplicate genes and NA's
  filter(dataset == "HUT set") %>%
  select(-dataset) %>%
  # rerank the atlas ranking and rank the fstats
  mutate(
    hrt_rank = dense_rank(hrt_rank),
    fstat_rank = dense_rank(fstat_rpkm),
    sds_rank = dense_rank(sds_rpkm),
    l2_rank = dense_rank(sqrt(hrt_rank^2 + fstat_rank^2 + sds_rank^2)),
    geom_rank = dense_rank(sqrt(hrt_rank * fstat_rank * sds_rank)),
  ) %>%
  # use the L2 normed ranking
  arrange(l2_rank)

# select top 4 genes
uf_finders <-
  uf_hs %$%
  external_gene_name %>%
  unique %>%
  head(4)

# look at the HUT set genes
uf_hs %>% {
  downtable(
    .,
    filterable = T,
    rowStyle = function(index)
      if (.$external_gene_name[index] %in% uf_finders) list(fontWeight = "bold")
    )
  }

```

Boxplots of the top newfound biomarker counts over cycle phases (in our training dataset of UF samples) and the four selected housekeeper genes in the bottom:

```{r warning = F, message = F, fig.height=21}

all_box <-
  gene_mat %>%
  filter(external_gene_name %in% uf_hs$external_gene_name) %>%
  pivot_longer(where(is_bare_numeric), names_to = "samplename", values_to = "molecule_count") %>%
  left_join(pheno %>% filter(group == "UF"), by = "samplename") %>%
  ggplot(aes(molecule_count, external_gene_name, fill = cyclephase)) +
  geom_boxplot() +
  theme_thesis() +
  scale_x_log10(breaks = breaks, minor_breaks = minor_breaks) +
  xlab("Molecule counts log10") +
  ylab("UF housekeeper candidates") +
  scale_fill_manual(values = color_grouping_alt) +
  theme(
    panel.grid.minor = element_line(),
    panel.grid.major = element_line()
  )

select_box <-
  gene_mat %>%
  filter(external_gene_name %in% uf_finders) %>%
  pivot_longer(where(is_bare_numeric), names_to = "samplename", values_to = "molecule_count") %>%
  left_join(pheno %>% filter(group == "UF"), by = "samplename") %>%
  ggplot(aes(molecule_count, external_gene_name, fill = cyclephase)) +
  geom_boxplot() +
  theme_thesis() +
  scale_x_log10(breaks = breaks, minor_breaks = minor_breaks) +
  xlab("Molecule counts log10") +
  ylab("UF housekeepers") +
  scale_fill_manual(values = color_grouping_alt) +
  theme(
    panel.grid.minor = element_line(),
    panel.grid.major = element_line()
  )


all_box / select_box + plot_layout(ncol = 1, heights = c(7, 1))

# ggsave("./figures/box_ted.svg", width = 6, height = 9)
```


# beREADY-EV testing

In this chapter we apply the new dataset and the set of biomarkers discovered in the last chapter to run the beREADY algorithm. Due to the ability of the `tacseqApp` to use drop-in replacements of the reference sets and the biomarker sets, then well try to hotwire the model to predict receptivity.

```{r warning = F, message = F}
## !RUNTHIS
add_coloring <- c(
  "LH2" = color_grouping_alt[["pre"]],
  "LH7" = color_grouping_alt[["rec"]],
  "LH12" = color_grouping_alt[["post"]],
  "Prolif" = color_grouping_alt[["pro"]],
  "Reference" = color_grouping_alt[["hrt"]]
  )
  
comb_pheno <- pheno

```


## UF+Biopsy biomarkers {.tabset}

Geneset of the intersection of UF and biopsy DE genes and TED genes. The Venn plot of showing the origin of selected genes:

```{r warning = F, message = F}
# !RUNTHIS
rep_or_not <- TRUE

## the hugo names for bio_genes
bio_genes_hugo <-
  annot %>%
  filter(ensembl_gene_id %in% bio_genes) %>%
  select("external_gene_name") %>%
  distinct(external_gene_name) %$%
  external_gene_name

uf_sig <- c(bio_genes_hugo, uf_finders)

# the gene "usage" table with added TED genes being default for all conditions
uf_dyn_genes <-
  bio_out_join %>%
  rename(ensembl_gene_id = locus) %>%
  geneid_converter(annot) %>%
  rename(locus = external_gene_name) %>%
  mutate(across(where(is_bare_numeric), as.logical)) %>%
  filter(locus %in% c(bio_genes_hugo, ted_genes)) %>%
  select(c("locus", contains("deseq_"))) %>%
  distinct(locus, .keep_all = T) %>%
  drop_na() %>%
  rename_with(~ str_remove(.x, "deseq_"), where(is.logical))

# emit a Venn diagram of the TED geneset in relation to other genes
uf_dyn_genes %>%
  mutate(TED = locus %in% ted_genes) %>%
  select(-locus) %>%
  ggplot(aes(A = pro_vs_pre, B = pre_vs_rec, C = rec_vs_post, D = TED)) +
  geom_venn(
    show_percentage = T, fill_alpha = 0.3
  ) + theme_thesis_venn()

uf_dyn_genes %<>%
  rename(
    pro = pro_vs_pre,
    pre = pre_vs_rec,
    post = rec_vs_post
  ) %>%
  mutate(across(where(is.logical), as.integer))
  
uf_dyn_genes %>% write_tsv(paste0(data_subfolder, "uf_bio_genes.tsv"))

```

```{r, child = "./analysis/test_samplesets.rmd"}
## use child rmarkdown document for running the testing repeatedly
```


## UF biomarkers {.tabset}

Large geneset of the UF DE genes and TED genes. The Venn plot of showing the origin of selected genes:

```{r warning = F, message = F}
# !RUNTHIS
rep_or_not <- TRUE

## the hugo names for bio_genes
uf_genes_hugo <-
  annot %>%
  filter(ensembl_gene_id %in% uf_genes) %>%
  select("external_gene_name") %>%
  distinct(external_gene_name) %$%
  external_gene_name

uf_sig <- c(uf_genes_hugo, uf_finders)


# the gene "usage" table with added TED genes being default for all conditions
uf_dyn_genes <-
  uf_out_join %>%
  rename(ensembl_gene_id = locus) %>%
  geneid_converter(annot) %>%
  rename(locus = external_gene_name) %>%
  mutate(across(where(is_bare_numeric), as.logical)) %>%
  filter(locus %in% c(uf_genes_hugo, ted_genes)) %>%
  select(c("locus", contains("deseq_"))) %>%
  distinct(locus, .keep_all = T) %>%
  drop_na() %>%
  rename_with(~ str_remove(.x, "deseq_"), where(is.logical))

# emit a Venn diagram of the TED geneset in relation to other genes
uf_dyn_genes %>%
  mutate(TED = locus %in% ted_genes) %>%
  select(-locus) %>%
  ggplot(aes(A = pro_vs_pre, B = pre_vs_rec, C = rec_vs_post, D = TED)) +
  geom_venn(
    show_percentage = T, fill_alpha = 0.3
  ) + theme_thesis_venn()

uf_dyn_genes %<>%
  rename(
    pro = pro_vs_pre,
    pre = pre_vs_rec,
    post = rec_vs_post
  ) %>%
  mutate(across(where(is.logical), as.integer))
  
uf_dyn_genes %>% write_tsv(paste0(data_subfolder, "uf_genes.tsv"))

```

```{r, child = "./analysis/test_samplesets.rmd"}
## use child rmarkdown document for running the testing repeatedly
```


# Adjusting the beREADY decision boundaries

Here we wrap the beREADY model with new decision boundaries.

First, let's see where are the sensible boundaries for our samples. We do observe a large discrepancy between post-receptive and late-receptive samples in the best models, so let's plot the continous "receptivity axis" with the original sample classes for every PCA subspace in our hierarchical model.

Run the beREADY-EV model again with the EV+Biopsy geneset, UF+Biopsy training samples and the EV housekeepers:

```{r warning = F, message = F}
rep_or_not <- TRUE

# the gene "usage" table with added TED genes being default for all conditions
uf_dyn_genes <-
  bio_out_join %>%
  rename(ensembl_gene_id = locus) %>%
  geneid_converter(annot) %>%
  rename(locus = external_gene_name) %>%
  mutate(across(where(is_bare_numeric), as.logical)) %>%
  filter(locus %in% c(bio_genes_hugo, ted_genes)) %>%
  select(c("locus", contains("deseq_"))) %>%
  distinct(locus, .keep_all = T) %>%
  drop_na() %>%
  rename_with(~ str_remove(.x, "deseq_"), where(is.logical))

uf_dyn_genes %<>%
  rename(
    pro = pro_vs_pre,
    pre = pre_vs_rec,
    post = rec_vs_post
  ) %>%
  mutate(across(where(is.logical), as.integer))

## with finders
the_gene_set <- c(uf_dyn_genes$locus, uf_finders)


## now format the data for beready and normalise the counts with hs
ref_dat <-
  raw_mat %>%
  rename(ensembl_gene_id = gene_id) %>%
  geneid_converter(annot) %>%
  rename(gene_id = external_gene_name) %>%
  format_for_beready(the_gene_set, pheno, uf_finders) %>%
  # remove samples which did not have any expression in the housekeeper genes and were NaN
  filter(if_all(where(is_bare_numeric), ~ !is.nan(.x) | !is.na(.x)))

test_dat <-
  raw_mat_test %>%
  rename(ensembl_gene_id = gene_id) %>%
  ## reannotate with HUGO
  geneid_converter(annot) %>%
  ## do the stupid name conversion again
  rename(gene_id = external_gene_name) %>%
  format_for_beready(the_gene_set, pheno_test, uf_finders)

## and run the model with the input reference data.
model_obj <-
  silent_beready_model(
    # the reference data
    ref_dat,
    # dynamic filtering geneset
    uf_dyn_genes,
    # and the data we want to evalate
    test_dat,
    is_dyn_filt = T
  )

```

Look on a simple line plot where those test samples where in relation to the "receptivity axes" of different model comparisons.

```{r warning = F, message = F, fig.height = 8}

rec_axes <-
  bind_rows(
    "all" = model_obj$dists_all %>% select(c("sample", starts_with("norm_"))),
    "pre_rec" = model_obj$dists_pre %>% select(c("sample", starts_with("norm_"))),
    "rec_post" = model_obj$dists_post %>% select(c("sample", starts_with("norm_"))),
    .id = "comparison"
  ) %>%
  rename(samplename = sample) %>%
  left_join(pheno_test, by = "samplename") %>%
  pivot_longer(starts_with("norm_"), values_to = "rec_axis", names_to = "rec_class")

rec_axes %>%
  drop_na(rec_axis) %>%
  ggplot(aes(rec_axis, rec_class, color = cyclephase)) +
  geom_point() + 
  facet_wrap(vars(comparison), ncol = 1, scales = "free") +
  scale_fill_manual(values = color_grouping_alt) +
  theme_minimal()

```

And the same plot for reported percentages:

```{r warning = F, message = F}

rec_axes <-
  model_obj$score_table %>%
  rename(samplename = `Sample Name`) %>%
  left_join(pheno_test, by = "samplename") %>%
  pivot_longer(ends_with("(%)"), values_to = "rec_axis", names_to = "rec_class")

rec_axes %>%
  drop_na(rec_axis) %>%
  ggplot(aes(rec_axis, rec_class, color = cyclephase)) +
  geom_point() + 
  scale_fill_manual(values = color_grouping_alt) +
  theme_minimal()

```

So, let's set the new boundary for the Post-receptive to Receptive to start from 60% and revisit the accuracy:

```{r warning = F, message = F}

model_obj$score_table %<>%
  mutate(`Predicted Group` = 
    case_when(
      `Pre-receptive (%)` > 60 ~ "Pre-receptive",
      `Pre-receptive (%)` > 40 ~ "Early-receptive",
      `Post-receptive (%)` > 60 ~ "Post-receptive",
      `Post-receptive (%)` > 40 ~ "Late-receptive",
      T ~ `Predicted Group`
      ))

res_table <-
  beready_results_formatter(
    model_obj$score_table,
    test_dat %>% select(c("sample", "group"))
  )

res_table %>%
  downtable(sortable = T, filterable = T)


vfold_split <-
  vfold_cv(
    ref_dat %>% select(-batch),
    v = 5,
    repeats = ifelse(rep_or_not, 20, 1),
    strata = group
    )

# run the CV runner in parallel and summarise the results
cv_results <-
  foreach::foreach(that_split = vfold_split$splits, .combine = "bind_rows", .inorder = F) %dopar% {
    beready_cv_fold(that_split, uf_dyn_genes)
  } %>%
  # select only the interesting values
  filter(str_detect(Metrics, "Acc")) %>%
  # pivot longer for grouping
  pivot_longer(where(is_bare_numeric), values_to = "Score", names_to = "Dataset") %>%
  # group by dataset type and summarise
  group_by(Dataset, Metrics) %>%
  # remove NAs in some cases to still get results
  summarise(`CV Accuracy` = mean(Score, na.rm = T)) %>%
  pivot_wider(values_from = "CV Accuracy", names_from = "Dataset") %>%
  mutate(Metrics = paste("CV", Metrics), .before = everything())

# get the summary stats from the testing set and combine it with crossval
beready_summary_stats(res_table) %>%
  bind_rows(cv_results) %>%
  mutate(across(where(is_bare_numeric), ~ round(.x, 4))) %>%
  downtable(sortable = T, defaultPageSize = 15)

```



# Session info

```{r warning = F, message = F}
sessionInfo()
```