---
title: "EV-biomarker study - xCell analysis"
author: Alvin Meltsov
date: "Compiled on `r format(Sys.time(), '%d/%m/%y')`"
output:
  rmdformats::robobook:
    number_sections: true
    code_folding: hide
    fig_width: 8
    fig_height: 6
---

# Prelude: Setup environment

## Load data and environment

Setup the environment and load dependencies, defined in external `back.R` script:

1. Load the dependencies, set up multithreading.
2. Set color groups and values.
3. Define convenience functions for preprocessing and visualisation.
4. Load and transform phenotype data and the data matrix.
5. Query BioMart for gene annotation and alternative names.
6. Load also important gene lists.

```{r warning = F, message = F}

## load the project dependencies
source("./deps.R")

## check if theres a R data object already, othewise run the data muncher 
# and save the preprocessed data into the RData
dataobj <- "./data/ev_data_filt.RData"
if (file.exists(dataobj)) {
  load(dataobj)
} else {
  ## load in the backend data muncher with arguments
  commandArgs <- function(...)
    list(
      switch_sample = "HUT23",
      remove_sample = c("HUT71_UF", "HUT10_UF", "HUT17_UF", "HUT1_UF")
      ) #  c("HUT71_UF", "HUT1_UF", "HUT01_UF", "HUT17_UF", "HUT10_UF"))
  invisible(source("./diff_run.R"))
  save.image(file = dataobj)
}

```

# How different are cell type profile in samples?

To estimate cell type profiles from bulk RNA-seq data, we use two commonplace methods:

1. XCell enrichment scores.
2. TODO: scProjection deconvolution based on single cell data. 

## XCell estimations

Despite looking like ratios limited 0..1, xCell emits two main metrics for cell type analysis - a p-value per prediction and the cell-type specific enrichment score.

To gain an overview of the dataset, we use interactive heatmaps to show significant scores and their values. Also, we perform PCA to visualise the differences in cell type profiles.

### HUT samples without filtering {.tabset}

Here we simply ran xCell on the reannotated dataset.

#### Original scores without filtering

```{r warning = F, message = F, fig.height = 16}

## read in the joined set dataobject
xcell_obj <- readRDS("./data/xcell_results.rds")$`HUT`
heatmap_annot <-
  # use here the "unfiltered" phenotype, as xCell was performed on all samples
  pheno_all %>%
  rename(grouping = group) %>%
  filter(samplename %in% colnames(xcell_obj$scores)) %>%
  arrange(colnames(xcell_obj$scores)) %>%
  column_to_rownames("samplename")


## TODO: set colors manually
htmap <-
  xcell_obj$scores %>%
    heatmaply(
      # plot_method = "ggplot",
      seriate = "OLO", # "mean", "OLO"
      scale = "none",
      dendrogram = "both",
      col_side_colors = heatmap_annot,
      # column_text_angle = 90,
      # row_side_palette = c(dim_color_grouping, "Input" = "#777777"),
      showticklabels = c(F, T)
    )

htmap

```

#### After filtering out cell types

Another heatmap where celltypes with less than quarter of significant (p<0.1) enrichment scores overall are excluded. Enrichment scores are also capped to 1 to keep better coloring for the lower range.

NB! "Smooth Muscle" was also removed as it does not provide much insight into different celltypes.

```{r warning = F, message = F, fig.height = 12}

xcell_enrich <-
  xcell_obj$scores %>%
  as_tibble(rownames = "celltype")

xcell_p <-
  xcell_obj$p_values %>%
  as_tibble(rownames = "celltype")

colnames(xcell_p) <- colnames(xcell_enrich)

xcell_sig <-
  xcell_p %>%
  mutate(across(where(is.numeric), ~ .x < 0.1)) %>%
  arrange(celltype)

xcell_filt <-
  xcell_enrich %>%
  select(colnames(xcell_sig)) %>%
  filter(celltype %in% xcell_sig$celltype) %>%
  # remove some extra cell types that don't make sense rn
  filter(!(celltype %in% c("Smooth muscle"))) %>%
  arrange(celltype) %>%
  column_to_rownames("celltype")

```

```{r warning = F, message = F, fig.height = 12}

## alternatively just exclude those tissue types that have low sig count
## without setting values to NA
## which breaks the hierarchical clustering algoritm
filt_sigs <-
  xcell_sig %>%
    rowwise() %>%
    transmute(
      celltype = celltype,
      nasum = sum(c_across(where(is.logical))),
    ) %>%
    filter(nasum / length(heatmap_annot$grouping) > 0.25)

htmap <-
  xcell_filt %>%
  as_tibble(rownames = "celltype") %>%
  filter(celltype %in% filt_sigs$celltype) %>%
  column_to_rownames("celltype") %>%
  mutate(across(where(is.numeric), ~ ifelse(.x > 1, 1, .x))) %>%
  heatmaply(
    # plot_method = "ggplot",
    seriate = "OLO", # "mean", "OLO"
    scale = "none",
    dendrogram = "both",
    col_side_colors = heatmap_annot,
    # column_text_angle = 90,
    # hide_colorbar = T,
    showticklabels = c(F, T)
  )

htmap

```

#### PCA of enrichment scores

Apply PCA dimensionality reduction for the xCell output. Filled out circles stand for UF samples.

```{r warning = F, message = F}

xcell_pre <-
  xcell_enrich %>%
  pivot_longer(
    where(is.numeric),
    names_to = "samplename",
    values_to = "xcell_enrich"
    ) %>%
  pivot_wider(
    id_cols = "samplename",
    names_from = "celltype",
    values_from = "xcell_enrich"
    )

expr_pca <-
  xcell_pre %>%
  recipe() %>%
  update_role(samplename, new_role = "id") %>%
  step_zv(all_numeric()) %>%
  step_normalize(all_numeric()) %>%
  step_pca(all_numeric(), num_comp = 3) %>%
  prep(strings_as_factors = F) %>%
  bake(new_data = xcell_pre) %>%
  left_join(heatmap_annot %>% as_tibble(rownames = "samplename"), by = "samplename")

```

```{r warning = F, message = F}

plot_ly() %>%
  add_trace(
    data = expr_pca %>% filter(grouping == "UF"),
    x = ~ PC1,
    y = ~ PC2,
    z = ~ PC3,
    type = "scatter3d",
    text = ~ samplename,
    color = ~ cyclephase,
    marker = list(size = 6),
    mode = "markers"
  ) %>%
  add_trace(
    data = expr_pca %>% filter(grouping == "biopsy"),
    x = ~ PC1,
    y = ~ PC2,
    z = ~ PC3,
    type = "scatter3d",
    text = ~ samplename,
    color = ~ cyclephase,
    marker = list(size = 6, symbol = "circle-open"),
    mode = "markers"
  )

```

### Large joined dataset run {.tabset}

For this option we will import data from the other previous studies:

* (Current data)
* Vigano EV (24),
* ART EV (after filtering 42),
* Endometrial and UF samples (20),
* GSE9838 (40) samples,
* cERtain brush biopsy samples (19),
* Blood extracellular vesicle long RNA profiles of human plasma by deep sequencing ("exoeasy1", 117),
* explored the plasma exoLRs profile of healthy individuals ("exoeasy2", 10).

#### Original scores without filtering

```{r warning = F, message = F, fig.height = 16}

## read in the joined set dataobject
xcell_obj <- readRDS("./data/xcell_results.rds")$`Joined set`
xcell_pheno <- readRDS("./data/xcell_results_pheno.rds")

heatmap_annot <-
  xcell_pheno %>%
    arrange(match(samplename, colnames(xcell_obj$scores))) %>%
    column_to_rownames("samplename")

num_col <- sum(
  length(unique(heatmap_annot$grouping)),
  length(unique(heatmap_annot$sampleset))
  )
ext_colors <-
  colorRampPalette(RColorBrewer::brewer.pal(8, "Set2"))(num_col) %>%
  setNames(c(unique(heatmap_annot$grouping), unique(heatmap_annot$sampleset)))

## TODO: manual colors
htmap <-
  xcell_obj$scores %>%
    heatmaply(
      # plot_method = "ggplot",
      seriate = "OLO", # "mean", "OLO"
      scale = "none",
      dendrogram = "both",
      col_side_colors = heatmap_annot,
      col_side_palette = ext_colors,
      # column_text_angle = 90,
      # row_side_palette = c(dim_color_grouping, "Input" = "#777777"),
      showticklabels = c(F, T)
    )

htmap

```

#### After filtering out

Another heatmap where celltypes with less than quarter of significant (p<0.1) enrichment scores overall are excluded. Enrichment scores are also capped to 1 to keep better coloring for the lower range.

```{r warning = F, message = F, fig.height = 12}

xcell_enrich <-
  xcell_obj$scores %>%
    as_tibble(rownames = "celltype")

xcell_p <-
  xcell_obj$p_values %>%
  as_tibble(rownames = "celltype")

colnames(xcell_p) <- colnames(xcell_enrich)

xcell_sig <-
  xcell_p %>%
  mutate(across(where(is.numeric), ~ .x < 0.1)) %>%
  arrange(celltype)

xcell_filt <-
  xcell_enrich %>%
  select(colnames(xcell_sig)) %>%
  filter(celltype %in% xcell_sig$celltype) %>%
  arrange(celltype) %>%
  column_to_rownames("celltype")

```

```{r warning = F, message = F, fig.height = 12}

## alternatively just exclude those tissue types that have low sig count
## without setting values to NA
## which breaks the hierarchical clustering algoritm
filt_sigs <-
  xcell_sig %>%
    rowwise() %>%
    transmute(
      celltype = celltype,
      nasum = sum(c_across(where(is.logical))),
    ) %>%
    filter(nasum / length(xcell_pheno$sampleset) > 0.25)

htmap <-
  xcell_filt %>%
  as_tibble(rownames = "celltype") %>%
  filter(celltype %in% filt_sigs$celltype) %>%
  column_to_rownames("celltype") %>%
  mutate(across(where(is.numeric), ~ ifelse(.x > 1, 1, .x))) %>%
  heatmaply(
    # plot_method = "ggplot",
    seriate = "OLO", # "mean", "OLO"
    scale = "none",
    dendrogram = "both",
    col_side_colors = heatmap_annot,
    col_side_palette = ext_colors,
    # column_text_angle = 90,
    # hide_colorbar = T,
    showticklabels = c(F, T)
  )
  ## don't show legend
  #layout(showlegend = F)

htmap

```

#### Selecting only new samples

The benefit of this plot compared to the "standalone run" one is that other related blood and endomerium sample help normalise the enrichment scores for the endometrial tissue.

```{r warning = F, message = F, fig.height = 12}

htmap <-
  xcell_filt %>%
  as_tibble(rownames = "celltype") %>%
  filter(celltype %in% filt_sigs$celltype) %>%
  # select current samples
  select(c("celltype", filter(xcell_pheno, str_detect(sampleset, "Combined HUT"))$samplename)) %>%
  column_to_rownames("celltype") %>%
  mutate(across(where(is.numeric), ~ ifelse(.x > 1, 1, .x))) %>%
  heatmaply(
    # plot_method = "ggplot",
    seriate = "OLO", # "mean", "OLO"
    scale = "none",
    dendrogram = "both",
    col_side_colors = 
      filter(xcell_pheno, str_detect(sampleset, "Combined HUT")) %>%
      select(-sampleset) %>%
      column_to_rownames("samplename"),
    # col_side_palette = ext_colors,
    # column_text_angle = 90,
    # hide_colorbar = T,
    showticklabels = c(F, T)
  )
  ## don't show legend
  #layout(showlegend = F)

htmap

```

#### Selecting only UF samples

```{r warning = F, message = F, fig.height = 12}

sel_samples <-
  xcell_pheno %>%
    filter(str_detect(grouping, "UF") & sampleset == "Combined HUT") %>%
    select(-sampleset)

htmap <-
  xcell_filt %>%
  as_tibble(rownames = "celltype") %>%
  filter(celltype %in% filt_sigs$celltype) %>%
  # select current samples
  select( c("celltype", sel_samples$samplename) ) %>%
  column_to_rownames("celltype") %>%
  mutate(across(where(is.numeric), ~ ifelse(.x > 1, 1, .x))) %>%
  heatmaply(
    # plot_method = "ggplot",
    seriate = "OLO", # "mean", "OLO"
    scale = "none",
    dendrogram = "both",
    col_side_colors = sel_samples %>% column_to_rownames("samplename"),
    # col_side_palette = ext_colors,
    # column_text_angle = 90,
    # hide_colorbar = T,
    showticklabels = c(F, T)
  )
  ## don't show legend
  #layout(showlegend = F)

htmap

```

#### Selecting only biopsy samples

```{r warning = F, message = F, fig.height = 12}

sel_samples <-
  xcell_pheno %>%
    filter(str_detect(grouping, "biopsy") & sampleset == "Combined HUT") %>%
    select(-sampleset)

htmap <-
  xcell_filt %>%
  as_tibble(rownames = "celltype") %>%
  filter(celltype %in% filt_sigs$celltype) %>%
  # select current samples
  select( c("celltype", sel_samples$samplename)) %>%
  column_to_rownames("celltype") %>%
  mutate(across(where(is.numeric), ~ ifelse(.x > 1, 1, .x))) %>%
  heatmaply(
    # plot_method = "ggplot",
    seriate = "OLO", # "mean", "OLO"
    scale = "none",
    dendrogram = "both",
    col_side_colors = sel_samples %>% column_to_rownames("samplename"),
    # col_side_palette = ext_colors,
    # column_text_angle = 90,
    # hide_colorbar = T,
    showticklabels = c(F, T)
  )
  ## don't show legend
  #layout(showlegend = F)

htmap

```


#### PCA of enrichment scores

Apply PCA dimensionality reduction for the xCell output.

```{r warning = F, message = F}

xcell_pre <-
  xcell_enrich %>%
  pivot_longer(
    where(is.numeric),
    names_to = "samplename",
    values_to = "xcell_enrich"
    ) %>%
  pivot_wider(
    id_cols = "samplename",
    names_from = "celltype",
    values_from = "xcell_enrich"
    )

expr_pca <-
  xcell_pre %>%
  recipe() %>%
  update_role(samplename, new_role = "id") %>%
  step_zv(all_numeric()) %>%
  step_normalize(all_numeric()) %>%
  step_pca(all_numeric(), num_comp = 3) %>%
  prep(strings_as_factors = F) %>%
  bake(new_data = xcell_pre) %>%
  left_join(xcell_pheno, by = "samplename")

```

```{r warning = F, message = F}

plot_ly() %>%
  add_trace(
    data = expr_pca %>% filter(!str_detect(grouping, "biopsy")),
    x = ~ PC1,
    y = ~ PC2,
    z = ~ PC3,
    type = "scatter3d",
    text = ~ samplename,
    color = ~ sampleset,
    marker = list(size = 6),
    mode = "markers"
  ) %>%
  add_trace(
    data = expr_pca %>% filter(str_detect(grouping, "biopsy")),
    x = ~ PC1,
    y = ~ PC2,
    z = ~ PC3,
    type = "scatter3d",
    text = ~ samplename,
    color = ~ sampleset,
    marker = list(size = 6, symbol = "circle-open"),
    mode = "markers"
  )

```

#### And UMAP

Empty circles represent biopsy samples, whereas filled out circles are UF samples.

```{r warning = F, message = F}

expr_umap <-
  xcell_pre %>%
  recipe() %>%
  update_role(samplename, new_role = "id") %>%
  step_zv(all_numeric()) %>%
  step_normalize(all_numeric()) %>%
  step_umap(all_numeric(), num_comp = 3) %>%
  prep(strings_as_factors = F) %>%
  bake(new_data = xcell_pre) %>%
  left_join(xcell_pheno, by = "samplename")

```

```{r warning = F, message = F}

plot_ly() %>%
  add_trace(
    data = expr_umap %>% filter(!str_detect(grouping, "biopsy")),
    x = ~ UMAP1,
    y = ~ UMAP2,
    z = ~ UMAP3,
    type = "scatter3d",
    text = ~ samplename,
    color = ~ sampleset,
    marker = list(size = 6),
    mode = "markers"
  ) %>%
  add_trace(
    data = expr_umap %>% filter(str_detect(grouping, "biopsy")),
    x = ~ UMAP1,
    y = ~ UMAP2,
    z = ~ UMAP3,
    type = "scatter3d",
    text = ~ samplename,
    color = ~ sampleset,
    marker = list(size = 6, symbol = "circle-open"),
    mode = "markers"
  )

```

### TODO: Static xCell heatmaps

```{r warning = F, message = F}

```


## TODO: scProjection

Here we apply the [scProjection](https://www.nature.com/articles/s41467-023-40744-6) model for the deconvolution. For the applidation, we need to use the combined HUTER/ER-atlas dataset for reference scRNA-seq data on cell types.


# Session info

```{r warning = F, message = F}
sessionInfo()
```